---
title: "Session2"
author: "Natalia Levshina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Creating "classic" and interactive D3 graphs of colexifications

You will need two packages: igraph and networkD3. Install them first, then load:

```{r message = FALSE}
library(igraph)
library(networkD3)
```
 
The dataset is called foot_graph.txt. It contains top frequent colexified concepts of FOOT.

```{r}
foot <- read.table("foot_graph.txt", header = T)
```

First, we make a traditional graph using functions from package igraph. We can see the vertices (nodes), the edges (links) and the weights (frequencies of colexifications).

```{r}
foot_graph <- graph_from_data_frame(foot, directed = FALSE)
V(foot_graph)#get the vertices (nodes)
E(foot_graph)#get the edges
E(foot_graph)$weight #get the weights (frequencies)
```

Of course, we can also plot the graph:


```{r}
plot(foot_graph)
```

The graph is easily customizable:

```{r}
plot(foot_graph, edge.width = E(foot_graph)$weight^0.4, edge.color = "lightgreen", vertex.label.cex = 1, vertex.size = 5, vertex.label.color = "black", vertex.label.family = "sans", vertex.color = "red", vertex.label.dist = 1.5)
```

This is informative, but sometimes it is practical to have an interactive graph. A popular tool is networkD3.

```{r}
foot_D3 <- igraph_to_networkD3(foot_graph)
forceNetwork(Links = foot_D3$links, Nodes = foot_D3$nodes, NodeID = 'name', Group = 'name', opacity = 0.9, Value = "value", linkDistance = 200, fontSize = 20, linkColour = "lightgreen", opacityNoHover = 0.9)
```

## Creating probabilistic maps of causatives

You will need several add-on packages: cluster and smacof for performing the MDS, fields, grDevices and googleVis for different visualizations.

```{r message = FALSE}
library(cluster)
library(smacof)
library(fields)
library(grDevices)
library(googleVis)
```
 
The dataset is causatives.txt. 

```{r}
causatives <- read.table("causatives.txt", header = T, sep = "\t", quote = "", stringsAsFactors = T)
str(causatives)
head(causatives)
```

The first step is to create a distance matrix using Gower distances. We only need the columns with the translations. 

```{r}
causatives.dist <- daisy(causatives[, 3:20])
summary(causatives.dist)
causatives.dist[1:3]
causatives[c(1, 2),]
causatives[c(1, 3),]
causatives[c(1, 4),]
```

Now we can represent these distances in a low-dimensional space using MDS. We will use metric MDS (the default). The default number of dimensions (ndim) is two.

```{r}
causatives.mds <- mds(causatives.dist) 
```

How many dimensions do we need?

```{r}
causatives.mds$stress
d1 <- mds(causatives.dist, ndim = 1)$stress 
d2 <- mds(causatives.dist, ndim = 2)$stress
d3 <- mds(causatives.dist, ndim = 3)$stress
d4 <- mds(causatives.dist, ndim = 4)$stress
d5 <- mds(causatives.dist, ndim = 5)$stress
plot(1:5, c(d1, d2, d3, d4, d5), type = "b", xlab = "n of dimensions", ylab = "Stress")
```

As we can see from the scree plot, there is no significant decrease in stress after two dimensions. So we will use the two-dimensional solution.


```{r}
plot(causatives.mds, "conf")
```

But what are these contexts? To interpret them, we can use interactive visualizations in googleVis. The resulting bubble chart will open in your default browser.

```{r}
text.df <- data.frame(Text = causatives$Text, Dim1 = causatives.mds$conf[, 1], Dim2 = causatives.mds$conf[, 2], ENG = causatives$ENG)
bubbles <- gvisBubbleChart(text.df, idvar = "Text", xvar = "Dim1", yvar = "Dim2", colorvar = "ENG", options = list(sizeAxis = '{maxSize: 10}', vAxis = '{minValue:-0.8, maxValue:0.8}', height = 500, width = 500, bubble="{textStyle:{color: 'none'}}"))
plot(bubbles)
```


Another common strategy is to plot the language-specific labels in order to compare the form-meaning mappings.

```{r}
plot(causatives.mds$conf, type = "n")
text(causatives.mds$conf, labels = causatives$ENG, cex = 0.5)
```

```{r}
plot(causatives.mds$conf, type = "n")
text(causatives.mds$conf, labels = causatives$FRA, cex = 0.5)
```

Let us now use different colours to represent the Romanian construction with face and the Italian construction with fare. We will add some jitter to avoid overlap.

```{r}
plot(causatives.mds$conf, type = "n", main = "fare vs. face")
points(jitter(causatives.mds$conf[causatives$ROM == "face",], amount = 0.1), col = "blue", pch = 16)
points(jitter(causatives.mds$conf[causatives$ITA == "fare",], amount = 0.1), col = "red", pch = 16)
legend("bottomright", legend = c("fare", "face"), col = c("red", "blue"), pch = 16, bty = "n")
```

Another strategy in situation of overlapping points is to use transparent colours. Below I use the function adjustcolor() from grDevices.

```{r}
plot(causatives.mds$conf, type = "n", main = "fare vs. face")
points(causatives.mds$conf[causatives$ROM == "face",], col = adjustcolor("blue", alpha.f = 0.5), pch = 16, cex = 1.5)
points(causatives.mds$conf[causatives$ITA == "fare",], col = adjustcolor("red", alpha.f = 0.5), pch = 16, cex = 1.5)
legend("bottomright", legend = c("fare", "face"), col = c("red", "blue"), pch = 16, bty = "n")
```

Sometimes it is useful to create a semantic area instead of isolated dots, with different densities, showing "peaks" with high expected frequency of a construction  and "valleys" with less likely semantic areas. We can use a geospatial technique called Kriging to interpolate likely values between the dots. The results can be easily represented as as heat maps, for example. The function Krig() is available in the package fields.

```{r}
y.ita <- ifelse(causatives$ITA == "fare", 1, 0)
y.fra <- ifelse(causatives$FRA == "faire", 1, 0)
y.spa <- ifelse(causatives$SPA == "hacer", 1, 0)
y.por <- ifelse(causatives$POR == "fazer", 1, 0)
y.rom <- ifelse(causatives$ROM == "face", 1, 0)
```

```{r}
Krig.rom <- Krig(causatives.mds$conf, y.rom, lambda = 0.05) #try different lambda values
Krig.ita <- Krig(causatives.mds$conf, y.ita, lambda = 0.05)
Krig.fra <- Krig(causatives.mds$conf, y.fra, lambda = 0.05)
Krig.spa <- Krig(causatives.mds$conf, y.spa, lambda = 0.05)
Krig.por <- Krig(causatives.mds$conf, y.por, lambda = 0.05)
```

Now the results can be visualized in a series of maps:

```{r}
par(mfrow = c(2, 3))
surface(Krig.ita, main = "ITA")
surface(Krig.fra, main = "FRA")
surface(Krig.spa, main = "SPA")
surface(Krig.por, main = "POR")
surface(Krig.rom, main = "ROM")
```
