---
title: "Session3"
author: "Natalia Levshina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading add-on packages

```{r message = FALSE}
#if needed, install the packages first
library(party)
library(Hmisc)
```


## Loading the dataset

```{r}
A2A_data <- read.table("A2A_data.txt", header = T, sep  = "\t", quote = "", comment = "", stringsAsFactors = T)
str(A2A_data)
```

First, we extracting the features for A and P separately from the initial dataset, and then merge them into one "long" dataset.

```{r}
data_A <- A2A_data[, c(1, 2, 4:9, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30:38, 41)]
data_P <- A2A_data[, c(1, 2, 4:9, 11, 13, 15, 17, 19, 21, 23, 25, 27, 29:38, 41)]
colnames(data_A) <- colnames(data_P) <- c("Language", "Filename", "Speaker_ID", "Text", "Gloss_Sentence", "Translation", "Word_Order", "Position_LexVerb",  "Wordform", "Gloss", "Case", "Person", "Semantics", "Structure", "Length_Syl", "Length_Words", "Discourse_Status", "Situation_Status", "Verb_Lemma", "Semantics_Verb", "Tense_Aspect", "Utterance_Type", "Clause", "Formal_Ambiguity", "Ditransitive", "Idiomatic", "FirstArg_Role", "Exclude")

data_A$Role <- as.factor(rep("A", nrow(data_A)))
data_P$Role <- as.factor(rep("P", nrow(data_P)))

data_long <- rbind(data_A, data_P)

data_long$Position <- rep("First", nrow(data_long))
data_long$Position[data_long$Role == "A"&data_long$FirstArg_Role == "P"] <- "Second"
data_long$Position[data_long$Role == "P"&data_long$FirstArg_Role == "A"] <- "Second"
data_long$Position <- as.factor(data_long$Position)

summary(data_long$Position)
```

We also need to get rid of noisy small categories, to facilitate machine learning.

```{r}
data_long$Situation_Status <- as.character(data_long$Situation_Status)
data_long$Situation_Status[data_long$Situation_Status %in% c("Present", "Absent", "Unclear")] <- NA
data_long$Situation_Status <- as.factor(data_long$Situation_Status)
summary(data_long$Situation_Status)

data_long$Discourse_Status <- as.character(data_long$Discourse_Status)
data_long$Discourse_Status[data_long$Discourse_Status %in% c("New", "Unclear")] <- NA
data_long$Discourse_Status <- as.factor(data_long$Discourse_Status)
summary(data_long$Discourse_Status)
```

## Exploring conditional inference trees

We create a dataset only with German contexts.

```{r}
ger_all <- data_long[data_long$Language == "German", 
c("Position", "Role", "Case", "Position_LexVerb", "Idiomatic", "Ditransitive", "Formal_Ambiguity", "Clause", "Tense_Aspect", "Utterance_Type", "Semantics_Verb", "Situation_Status", "Discourse_Status", "Length_Syl", "Length_Words","Structure", "Semantics", "Person", "Filename")]

ger_all <- ger_all[complete.cases(ger_all),]#611 obs.

ger_all$Structure <- factor(ger_all$Structure)
ger_all$Tense_Aspect <- factor(ger_all$Tense_Aspect)
ger_all$Filename <- factor(ger_all$Filename)
```

What does a conditional inference tree look like? We will use the default settings.

```{r}
mytree_ger <- ctree(Position ~ ., data = ger_all)
plot(mytree_ger)
```

Repeat the procedure for the other two languages.

```{r}
rus_all <- data_long[data_long$Language == "Russian", 
c("Position", "Role", "Case", "Position_LexVerb", "Idiomatic", "Ditransitive", "Formal_Ambiguity", "Clause", "Tense_Aspect", "Utterance_Type", "Semantics_Verb", "Situation_Status", "Discourse_Status", "Length_Syl", "Length_Words", "Structure", "Semantics", "Person", "Filename")]

rus_all <- rus_all[complete.cases(rus_all),]

rus_all$Structure <- factor(rus_all$Structure)
rus_all$Tense_Aspect <- factor(rus_all$Tense_Aspect)
rus_all$Filename <- factor(rus_all$Filename)

mytree_rus <- ctree(Position ~ ., data = rus_all)
plot(mytree_rus)


tag_all <- data_long[data_long$Language == "Tagalog", 
c("Position", "Role", "Case", "Position_LexVerb", "Idiomatic", "Ditransitive", "Formal_Ambiguity", "Clause", "Tense_Aspect", "Utterance_Type", "Semantics_Verb", "Situation_Status", "Discourse_Status", "Length_Syl", "Length_Words", "Structure", "Semantics", "Person", "Filename")]

tag_all <- tag_all[complete.cases(tag_all),]

tag_all$Structure <- factor(tag_all$Structure)
tag_all$Tense_Aspect <- factor(tag_all$Tense_Aspect)
tag_all$Filename <- factor(tag_all$Filename)

mytree_tag <- ctree(Position ~ ., data = tag_all)
plot(mytree_tag)
```

## Growing random forests

Next, we can grow a 1000 trees forest using mtry = 5. 

```{r}
myforest_ger_all_mtry5 <- cforest(Position ~ ., data = ger_all, controls = cforest_unbiased(ntree = 1000, mtry = 5))
importance_ger_all_mtry5 <- varimp(myforest_ger_all_mtry5, conditional = T) #this can take a while

cutoff <- abs(min(importance_ger_all_mtry5))
dotchart(sort(importance_ger_all_mtry5), main = "German: cforest, Variable Importance, mtry = 5")
abline(v = cutoff, lty = 2, col = "blue")
```

Exercise: try other mtry values (e.g., 10 and 15). Compare the results.

A random forests for Russian:

```{r}
myforest_rus_all_mtry5 <- cforest(Position ~ ., data = rus_all, controls = cforest_unbiased(ntree = 1000, mtry = 5))
importance_rus_all_mtry5 <- varimp(myforest_rus_all_mtry5, conditional = T)

cutoff <- abs(min(importance_rus_all_mtry5))
dotchart(sort(importance_rus_all_mtry5), main = "Russian: cforest, Variable Importance, mtry = 5")
abline(v = cutoff, lty = 2, col = "blue")
```

A random forest for Tagalog:

```{r}
myforest_tag_all_mtry5 <- cforest(Position ~ ., data = tag_all, controls = cforest_unbiased(ntree = 1000, mtry = 5))
importance_tag_all_mtry5 <- varimp(myforest_tag_all_mtry5, conditional = T)

cutoff <- abs(min(importance_tag_all_mtry5))
dotchart(sort(importance_tag_all_mtry5), main = "Tagalog cforest, Variable Importance, mtry = 5")
abline(v = cutoff, lty = 2, col = "blue")
```

Try other mrty values and compare the results.
What can you conclude?

## Evaluating the models' performance

Computing Out-of-Bag (OOB) C-score for German:

```{r}
pred_oob_ger <- predict(myforest_ger_all_mtry5, type = "prob", OOB = TRUE)
prob_oob_ger <- unlist(pred_oob_ger)[c(FALSE, TRUE)]
somers2(prob_oob_ger, as.numeric(ger_all$Position) - 1)
```

Compute the OOB accuracy score:

```{r}
pred_oob_cat_ger <- predict(myforest_ger_all_mtry5, type = "response", OOB = TRUE)
table(pred_oob_cat_ger, ger_all$Position)
```

Compute the accuracy score by dividing correct predictions by the total number of observations.
Exercise: try the same for Russian and Tagalog.